\pkg{hesim}: Health Economic Simulation Modeling and Decision Analysis

### 2020-12-20 manuscript.Rnw
Complete draft of illustrative example

%% -- Individual continuous time state transition model ------------------------
\subsection{Individual continuous time state transition model}
We setup the model by defining the model structure, target population, and treatment strategies of interest. The transitions are characterized by a matrix where each element denotes a transition from a row to a column. If a transition is not possible it is marked with \code{NA}; otherwise, an integer denotes the transition number.

@@

%% -- Partitioned survival model -----------------------------------------------
\subsection{Partitioned survival model}
In the previous section multi-state models were fit to individual patient data. Partitioned survival analysis is an alternative approach that is frequently used for economic evaluations of therapies in oncology. Although there is increasing evidence that PSM are more biased than multi-state models \citep{woods2020partitioned}, PSMs are easier to parameterize with aggregate level data. Techniques for fitting multi-state models with aggregate data require more constraints and are less established \citep{price2011parameterization, pahuta2019technique, jansen2020multi}. PSMs, in contrast, consist of separate independent survival models that can be estimated with aggregate level data by reconstructing individual patient data from digitized survival curves \citep{guyot2012enhanced}.

@@

Finally, the complete partitioned survial model is constructed using the survival, utility, and cost mdoels.

@@

\subsection{Cohort disctete time state transition model} \label{sec:cdtstm-example}

\subsubsection{Parameterization}

<<cdtstm_fit_soc, warning = FALSE>>=
library("msm")
qinit <- matrix(0, nrow = 3, ncol = 3)
qinit[1, 2] <-  0.28
qinit[1, 3] <-  0.013
qinit[2, 3] <-  0.10
msm_fit <- msm(state_id ~ time, subject = patient_id, 
               data = onc3p[strategy_name == "SOC"],
               exacttimes = FALSE,
               covariates = ~ age + female,
               qmatrix = qinit, gen.inits = FALSE)
@

<<msm_exact, eval = FALSE, message = FALSE>>=
# msm with exact transition times
tmat2 <- tmat; tmat2[is.na(tmat2)] <- 0
msm_efit <- msm(state_id ~ time, subject = patient_id, 
                data = onc3p[strategy_name == "SOC"],
                qmatrix = tmat2, exacttimes = TRUE)

@@

<<cdtstm_params>>=
params2 <- list(
  lrr_12_est = c(soc = log(1), new1 = log(.80), new2 = log(.71)),
  lrr_12_se = c(soc = 0, new1 = .03, new2 = .04),
  lrr_13_est = c(soc = log(1), new1 = log(.90), new2 = log(.85)),
  lrr_13_se = c(soc = 0, new1 = .02, new2 = .03),
  u_mean = c(s1 = .9, s2 = .75, s3 = .05),
  u_se = c(s1 = .02, s2 = .03, s3 = .05)
)

params2_def <- define_rng({
  list(
    rr_12 = lognormal_rng(lrr_12_est, lrr_12_se),
    rr_13 = lognormal_rng(lrr_13_est, lrr_13_se),
    u = beta_rng(mean = u_mean, sd = u_se)
  )}, n = n_samples)

params2_rng <- eval_rng(params2_def, params2)
@

<<qmat_soc>>=
transmod_data <- survmods_data
qmat_soc <- qmatrix(msm_fit, newdata = transmod_data[strategy_name == "SOC"],
                    uncertainty = "normal", n = n_samples)
qmat_soc[,, 1]
@

@@

@@

@@

<<rr>>=
x_rr <- as.matrix(transmod_data[, .(soc, new1, new2)])
rr <- cbind(xbeta(x_rr, params2_rng$rr_12),
            xbeta(x_rr, params2_rng$rr_13))
@

<<apply_rr>>=
pmat <- apply_rr(pmat_soc, rr = rr,
                 index = list(c(1, 2), c(1, 3)))
@

<<tparams_transprobs>>=
tprobs <- tparams_transprobs(pmat, tpmat_id)
@

\subsubsection{Simulation}
<<dtstm_transmod>>=
transmod <- CohortDtstmTrans$new(params = tprobs, cycle_length = cycle_len)
@

@@

@@

@@

<<cdtstm_icer>>=
ce_sim_cdtstm <- cdtstm$summarize()
cea_pw_cdtstm <- cea_pw(ce_sim_cdtstm, comparator = 1, 
                        dr_costs = .03, dr_qalys = .03)
icer_tbl(cea_pw_cdtstm, output = "data.table")
@

>>>>

%% -- Individual continuous time state transition model ------------------------
\subsection{Individual continuous time state transition model} \label{sec:ictstm-example}
We setup the model by defining the model structure, target population, and treatment strategies of interest. The transitions are characterized by a matrix where each element denotes a transition from a row to a column. If a transition is not possible it is marked with \code{NA}; otherwise, an integer denotes the transition number.

@@

%% -- Partitioned survival model -----------------------------------------------
\subsection{Partitioned survival model} \label{sec:psm-example}
In the previous section multi-state models were fit to individual patient data. Partitioned survival analysis is an alternative approach that is frequently used for economic evaluations of therapies in oncology. Although there is increasing evidence that PSM are more biased than multi-state models \citep{woods2020partitioned}, PSMs are easier to parameterize with aggregate level data. Techniques for fitting multi-state models with aggregate data require more constraints and are less established \citep{price2011parameterization, pahuta2019technique, jansen2020multi}. PSMs, in contrast, consist of separate independent survival models that can be estimated with aggregate level data by reconstructing individual patient data from digitized survival curves \citep{guyot2012enhanced}.

@@

Finally, the complete partitioned survial model is constructed using the survival, utility, and cost models

@@

%% -- Cohort discrete time state transition model ------------------------------
\subsection{Cohort disctete time state transition model} \label{sec:cdtstm-example}
The third model type supported by \pkg{hesim} is the cDTSTM. Although it is commonly used in health economics, the approaches taken in Section~\ref{sec:ictstm-example} and \ref{sec:psm-example} make better use of multi-state data if the time to event outcomes are continuously observed. However, in some cases panel data is available where observations are only collected at arbitrary times. This may, for instance, occur in routine practice oncology data when progression of disease is not continuously evaluated. To facilitate analysis of this type of data, we have converted the \code{onc3} dataset from above into panel form.

<<cdtstm_data, warning = FALSE>>=
head(onc3p)
@

\subsubsection{Parameterization}
As in Section~\ref{sec:psm-example}, we consider a case where patient level data is available for SOC but aggregate level data is used to estimate the relative treatment effects. When exact transition times are not observed, the survival models estimated in Section~\ref{sec:ictstm-example} are inappropriate. The \pkg{msm} package, can, however, be used to estimate models from such data. The loss of information does come at a price as only exponential and piecewise exponential models are possible. We illustrate by fitting an exponential multi-state model for SOC using \pkg{msm}. (Note that if exact times were observed and we set the argument \code{exacttimes = TRUE}, then the parameter estimates would be the same as an exponential model fit using \code{flexsurvreg()}.)

<<cdtstm_fit_soc, warning = FALSE, message = FALSE>>=
library("msm")
qinit <- matrix(0, nrow = 3, ncol = 3)
qinit[1, 2] <-  0.28; qinit[1, 3] <-  0.013; qinit[2, 3] <-  0.10
msm_fit <- msm(state_id ~ time, subject = patient_id, 
               data = onc3p[strategy_name == "SOC"],
               exacttimes = FALSE,
               covariates = ~ age + female,
               qmatrix = qinit, gen.inits = FALSE)
@

<<msm_exact, include = FALSE, message = FALSE>>=
# msm with exact transition times
tmat2 <- tmat; tmat2[is.na(tmat2)] <- 0
msm_efit <- msm(state_id ~ time, subject = patient_id, 
                data = onc3p[strategy_name == "SOC"],
                qmatrix = tmat2, exacttimes = TRUE)

@@

We assume that the relative treatment effects were estimated as relative risks. Either a posterior distribution from a Bayesian analysis or parameters of a parametric distribution can be used. We use the latter approach here and assume that the relative risk is approximately normally distributed on the log scale. \pkg{hesim} provides the \code{define_rng()} and \code{eval_rng()} functions for defining and evaluating random number generators for a PSA. While standard functions in \proglang{R} could be used to draw random deviates from probability distributions instead, these functions can be more convenient for health economic modeling. Notable advantages include only needing to specify the number of random samples once, being able to pass moments of distributions as arguments (as in \code{stateval_tbl()}), and guaranteeing a common output (either a vector or a two-dimensional tabular object, with dimensions determined by the number or random samples).

<<cdtstm_params>>=
params_rr <- list(
  lrr_12_est = c(soc = log(1), new1 = log(.80), new2 = log(.71)),
  lrr_12_se = c(soc = 0, new1 = .03, new2 = .04),
  lrr_13_est = c(soc = log(1), new1 = log(.90), new2 = log(.85)),
  lrr_13_se = c(soc = 0, new1 = .02, new2 = .03)
)

params_rr_def <- define_rng({
  list(
    rr_12 = lognormal_rng(lrr_12_est, lrr_12_se),
    rr_13 = lognormal_rng(lrr_13_est, lrr_13_se)
  )}, n = n_samples)
params_rr_rng <- eval_rng(params_rr_def, params_rr)
@

In the prior two sections, we constructed the disease model using parameter (i.e., \code{params_surv}) objects. Here, we will use a transformed parameter (i.e., \code{tparams_transprobs}) object, which is a very flexible way to construct the transition probability matrices for a cDTSTM. A \code{tparams_transprobs} object requires a 3-dimensional array of transition matrices, sorted by PSA sample, treatment strategy, patient, and optionally (in a time-inhomogeneous model) time interval.

We start by predicting transition intensity matrices for SOC from the exponential multi-state model using the same input data used for the survival models in the PSM example. A prediction is made for each representative patient and samples of the transition intensity matrices are drawn for the PSA using the asymptotic normal approximation. The array contains $B \times N_p$ matrices where $B$ is the number of PSA samples and $N_p$ is the number of patients.

<<qmat_soc>>=
transmod_data <- survmods_data
qmat_soc <- qmatrix(msm_fit, newdata = transmod_data[strategy_name == "SOC"],
                    uncertainty = "normal", n = n_samples)
dim(qmat_soc)
qmat_soc[,, 1]
@

Transition probability matrices of arbitrary duration are derived from the transition intensity matrices with the matrix exponential. We use a cycle length of 3 months.

@@

Identifiers for each matrix in the array can then be generated with \code{tpmatrix_id()}.

@@

Relative risks can be predicted for each row of the input data and each sample of the coefficients for the PSA using matrix multiplication. In particular, let $x$ be a $N \times k$ matrix where $k$ is the number of predictors and $\beta$ be a $B \times k$ matrix where each row contains a given sample for the PSA. The function \code{xbeta} produces a $N \times B$ matrix of predicted relative risks and then ``flattens'' the matrix to create a single vector of length $B \times N$.

@@

We use this function to predict distributions of relative risks (for the transitions stable $\rightarrow$ progression and stable $\rightarrow$ death) as a function of the three treatment strategies.

<<rr>>=
x_rr <- as.matrix(transmod_data[, .(soc, new1, new2)])
rr <- cbind(xbeta(x_rr, params_rr_rng$rr_12),
            xbeta(x_rr, params_rr_rng$rr_13))
@

Relative risks are applied to the elements (stable $\rightarrow$ progression (1, 2) and stable $\rightarrow$ death (1, 3)) of each transition probability matrix for SOC with \code{apply_rr()}. The result is an array of matrices for each PSA sample, treatment strategy,  and representative patient, as desired. The \code{tparams_transprobs} object can then be created by simply combining the transition probability matrices and the identification dataset.

<<apply_rr>>=
pmat <- apply_rr(pmat_soc, rr = rr,
                 index = list(c(1, 2), c(1, 3)))
tprobs <- tparams_transprobs(pmat, tpmat_id)
@

\subsubsection{Simulation}
One the model has been parameterized, the simulation is straightforward. The transition model depends only on the transition probability matrices and the cycle length,

<<dtstm_transmod>>=
transmod <- CohortDtstmTrans$new(params = tprobs, cycle_length = cycle_len)
@

and the complete economic model is again instantiated from the transition, utility, and cost submodels.

@@

We simulate state probabilities for 120 cycles (i.e., 30 years).

@@

QALYs and costs are again simulated using a 3 percent discount rate

@@

and an ICER is computed.

<<cdtstm_icer>>=
ce_sim_cdtstm <- cdtstm$summarize()
cea_pw_cdtstm <- cea_pw(ce_sim_cdtstm, comparator = 1, 
                        dr_costs = .03, dr_qalys = .03)
icer_tbl(cea_pw_cdtstm, output = "data.table", k = 100000)
@
